\textbf{Effective application of machine learning to microbiome based classification problems}

### 2019-10-14 submission/manuscript.tex
Jenna's edits 3rd round

Machine learning (ML) modeling of the human microbiome has the potential
to identify microbial biomarkers and aid in diagnosis of many diseases
such as inflammatory bowel disease, diabetes, and colorectal cancer.
Progress has been made towards developing ML models that predict health
outcomes using bacterial abundances, but inconsistent adoption of
training and evaluation methods call the validity of these models into
question. Furthermore, there appears to be a preference by many
researchers to favor increased model complexity over interpretability.
To overcome these challenges, we trained seven models that used fecal
16S rRNA sequence data to predict the presence of colonic screen
relevant neoplasias (SRNs; n=490 patients, 261 controls and 229 cases).
We developed a reusable open-source pipeline to train, validate and
interpret the models. To show the effect of model selection, we assessed
the predictive performance, interpretability, and training time of
L2-regularized logistic regression, L1 and L2-regularized support vector
machines (SVM) with linear and radial basis function kernels, decision
trees, random forest, and gradient boosted trees (XGBoost). The random
forest model performed best at detecting SRNs with an AUROC of 0.695
{[}IQR 0.651-0.739{]} but was slow to train (83.2 h) and not immidiately
interpretable. Despite its simplicity, L2-regularized logistic
regression followed random forest in predictive performance with an
AUROC of 0.680 {[}IQR 0.625-0.735{]}, trained faster (12 min), and
yielded a model that was inherently interpretable. Our analysis
highlights the importance of choosing an ML approach should be chosen
based on the goal of the study, as the choice will inform expectations
of performance and interpretability.

@@

\textbf{Model selection and pipeline construction} We established a
reusable ML pipeline for model selection and evaluation, focusing on
seven different commonly used supervised learning algorithms {[}Figure
1{]}.

@@

\textbf{Predictive performance and generalizability of the seven
models.} We evaluated the predictive performance of the seven models to
classify individuals as having normal colons or SRNs {[}Figure 2{]}. The
random forest model had a median AUROC value of 0.695 {[}IQR
0.650-0.739{]} which was higher than the other models for detecting
SRNs. However, we determined that the random forest model performance
was not statistically significantly higher (p = 0.5) when we
investigated the distribution of differences between random forest and
L2-regularized logistic regression AUROC values for each data-split
(Figure S3). Similarly, L2-regularized logistic regression, XGBoost,
L2-regularized SVM with linear and radial basis function kernel AUROC
values were not significantly different from one another and had median
AUROC values of 0.680 {[}IQR 0.639-0.750{]}, 0.679 {[}IQR
0.643-0.746{]}, 0.678 {[}IQR 0.639-0.750{]} and 0.668 {[}IQR
0.639-0.750{]}, respectively. L1-regularized SVM with linear kernel and
decision tree had significantly lower AUROC values than the other ML
models with median AUROC of 0.650 {[}IQR 0.629-0.760{]} and 0.601 {[}IQR
0.636-0.753{]}, respectively {[}Figure 2{]}. Interestingly, these
results demonstrate that the most complex model (XGBoost) did not have
the best performance and that the most interpretable models
(L2-regularized logistic regression and L2-regularized SVM with linear
kernel) performed nearly as well non-linear models.

@@

Second, to analyze non-linear models we interpreted the feature
importance using permutation importance (36). Whereas the absolute
feature weights were determined from the trained models, here we
measured importance using the held-out test data. Permutation importance
analysis is a post hoc explanation of the model, in which we randomly
permuted groups of perfectly correlated features together and other
features individually across the two groups in the held-out test data.
We then calculated how much the predictive performance of the model
(i.e, testing AUROC values) decreased when each OTU or group of OTUs was
randomly permuted. We ranked the OTUs based on how much the median
testing AUROC decreased when it was permuted; the OTU with the largest
decrease ranked highest {[}Figure 4{]}. Among the twenty OTUs with the
largest impact, there was only one OTU (OTU 822) that was shared among
all of the models; however, we found three OTUs (OTUs 58, 110, 367) that
were important in each of the tree-based models. Similarly, the random
forest and XGBoost models, shared four of the most important OTUs (OTUs
2, 12, 361, 477). Permutation analysis results also revealed that with
the exception of the decision tree model, removal of any individual OTU
had minimal impact on model performance. For example, if OTU 367 was
permuted across the samples in the decision tree model, the median AUROC
dropped from 0.601 to 0.525. In contrast, if the same OTU was permuted
in the random forest model, the AUROC only dropped from 0.695 to 0.680.
Although permutation analysis allowed us to gauge the importance of an
OTU, the analysis was post hoc (i.e., done using the test data).

@@

Our framework provides a reproducible structure to investigators wanting
to train, evaluate, and interpret their own ML models to generate
hypotheses regarding which OTUs might be biologically relevant. However,
deploying microbiome-based models to make clinical diagnoses or
predictions is a significantly more challenging and distinct
undertaking. For example, we currently lack standardized methods to
collect patient samples, generate sequence data, and report clinical
data. We are also challenged by the practical constraints of OTU-based
approaches. The de novo algorithms commonly in use are slow, require
considerable memory, and result in different OTU assignments as new data
are added (41). Finally, we also need independent validation cohorts to
test the performance of a diagnostic model. To realize the potential for
using ML approaches with microbiome data, it is necessary that we direct
our efforts to overcome these challenges.

@@

\textbf{Data collection and study population.} The original stool
samples described in our analysis were obtained from patients recruited
by Great Lakes-New England Early Detection Research Network (5). Stool
samples were provided by adults who were undergoing a scheduled
screening or surveillance colonoscopy. Participants were recruited from:
Toronto (ON, Canada), Boston (MA, USA), Houston (TX, USA), and Ann Arbor
(MI, USA). Patients' colonic health was visually assessed by colonoscopy
with bowel preparation and tissue histopathology of all resected
lesions. We assigned patients into two classes: those with normal colons
and those with screen relevant neoplasias (SRNs). The normal class
included patients with normal colons or non-advanced adenomas whereas
the SRN class included patients with advanced adenomas or carcinomas
(42). Patients with an adenoma greater than 1 cm, more than three
adenomas of any size, or an adenoma with villous histology were
classified as having advanced adenomas. There were 172 patients with
normal colonoscopies, 198 with adenomas, and 120 with carcinomas. Of the
198 adenomas, 109 were identified as advanced adenomas. Together 261
patients were classified as normal and 229 patients were classified as
having a SRN.

\textbf{16S rRNA gene sequencing data.} Stool samples provided by the
patients were used for 16S rRNA gene sequencing to measure bacterial
population abundances. The sequence data used in our analyses were
originally generated by Baxter et al. (available through NCBI Sequence
Read Archive {[}SRP062005{]}, 2015). The OTU abundance table was
generated by Sze et al (43), who processed the 16S rRNA sequences in
mothur (v1.39.3) using the default quality filtering methods,
identifying and removing chimeric sequences using VSEARCH, and assigning
to OTUs at 97\% similarity using the OptiClust algorithm (41, 44, 45);
(\url{https://github.com/SchlossLab/Sze_CRCMetaAnalysis_mBio_2018/blob/master/data/process/baxter/baxter.0.03.subsample.shared}).
These OTU abundances were the features we used to predict colorectal
health of the patients. There were 6920 OTUs. OTU abundances were
subsampled to the size of the smallest sample and normalized across
samples such that the highest abundance of each OTU would be 1 and
lowest would be 0.

@@

For hyperparameter selection, we started with a granular grid search.
Then we narrowed and fine-tuned the range of each hyperparameter. A grid
search was needed to avoid large variation in prediction performance.
Recently developed tools such as Hyperband help researchers with
hyperparameter selection that can be incorporated to microbiome studies
(46). For L2-regularized logistic regression, L1 and L2-regularized SVM
with linear and radial basis function kernels, we tuned the cost
hyperparameter, which controls the regularization strength, where
smaller values specify stronger regularization. For SVM with radial
basis function kernel, we also tuned the sigma hyperparameter which
determines the reach of a single training instance where for a high
value of sigma, the SVM decision boundary will be dependent on the
points that are closest to the decision boundary. For the decision tree
model, we tuned the depth of the tree where the deeper the tree, the
more splits it has. For random forest, we tuned the number of features
to consider when looking for the best tree split. For XGBoost, we tuned
the learning rate and the fraction of samples used for fitting the
individual base learners.

@@

\textbf{Statistical analysis workflow.} Data summaries, statistical
analysis, and data visualizations were performed using R (v.3.5.0) with
the tidyverse package (v.1.2.1). We compared the performance of the
models by taking the difference of AUROC values from the same datasplit.
We plotted a histogram that showed the AUROC differences for 100
datasplits and calculated the empirical p-value of the histogram (2 x
min(\% of AUROC differences \(\geq\) 0, \% of AUROC differences \(\leq\)
0) for double tail event (e.g., Figure S3).

@@

\textbf{Acknowlegements.} We thank all the study participants of Great
Lakes-New England Early Detection Research Network. We would like to
thank the members of the Schloss lab for their valuable feedback during
the formulation of the manuscript. Salary support for M.R came from NIH
grant 1R01CA215574. Salary support for P.D.S. came from NIH grants
P30DK034933 and 1R01CA215574.

@@

\textbf{Figure 2. Generalization and classification performance of ML
models using AUROC values of all cross validation and testing
performances.} The median AUROC for diagnosing individuals with SRN
using bacterial abundances was higher than chance (depicted by
horizontal line at 0.50) for all the ML models. The predictive
performance of random forest model was higher than other ML models
though not significantly (p=0.5). L2-regularized logistic regression,
XGBoost, L2-regularized SVM with linear and radial basis function kernel
performances were not significantly different from one another. The
boxplot shows quartiles at the box ends and the median as the horizontal
line in the box. The whiskers show the farthest points that were not
outliers. Outliers were defined as those data points that are not within
1.5 times the interquartile ranges. \newpage
\includegraphics[height=17cm, width=12cm]{Figure_3.png}

\textbf{Figure 3. Interpretation of the linear ML models.} The ranks of
absolute feature weights of (A) L2-regularized logistic regression, (B)
L1-regularized SVM with linear kernel, and (C) L2-regularized SVM with
linear kernel were ranked from highest rank, 1, to lowest rank, 100, for
each data-split. The feature ranks of the 20 highest ranked OTUs based
on their median ranks (median shown in black) are reported here. OTUs
that are associated with classifying a subject as being healthy had
negative signs and were shown in blue. OTUs that are associated with
classifying a subject having an SRN had positive signs and were shown in
red. Some of the same OTUs were identified as important in all of the
linear models. \newpage
\includegraphics{Figure_4.png}

\textbf{Figure 4. Interpretation of the non-linear ML models.} (A) SVM
with radial basis kernel, (B) decision tree, (C) random forest, and (D)
XGBoost feature importances were explained using permutation importance
on the held-out test data set. The gray rectangle and the dashed line
show the IQR range and median of the base testing AUROC without any
permutation. The colors of the box plots represent the OTUs that were
shared among the different models; yellow were OTUs that were shared
among all the non-linear models, salmon were OTUs that were shared among
the tree-based models, green were the OTUs shared among SVM with radial
basis kernel, decision tree and XGBoost, pink were the OTUs shared among
SVM with radial basis kernel and XGBoost only, red were the OTUs shared
among random forest and XGBoost only and blue were the OTUs shared among
decision tree and random forest only. For all of the tree-based models,
a \emph{Peptostreptococcus} species (OTU00367) had the largest impact on
predictive performance. \newpage
\includegraphics{Figure_5.png}

@@

\textbf{Figure S1. Hyperparameter setting performances for linear
models.} (A) L2 logistic regression, (B) L1 SVM with linear kernel, and
(C) L2 SVM with linear kernel mean cross-validation AUROC values when
different hyperparameters were used in training the model. The stars
represent the highest performing hyperparameter setting for each model.
\newpage
\includegraphics[height=30cm, width=15cm]{Figure_S2.png}

\textbf{Figure S2. Hyperparameter setting performances for non-linear
models.} (A) Decision tree, (B) Random forest, (C) SVM with radial basis
kernel, and (D) XGBoost mean cross-validation AUROC values when
different hyperparameters were used in training the model. The stars
represent the highest performing hyperparameter setting for the models.
\newpage
\includegraphics[height=17.5cm, width=13cm]{Figure_S3.png}

@@

\textbf{Figure S4. Classification performance of ML models across cross
validation when trained on a subset of the dataset.} (A) L2-regularized
logistic regression and (B) Random forest models were trained using the
original study design with 490 subjects and subsets of it with 15, 30,
60, 120, and 245 subjects. The range among the cross-validation AUROC
values within both models at lower sample sizes were much larger than
when the full collection of samples was used to train and validate the
models, but included the ranges observed with the more complete
datasets. \newpage
\includegraphics[height=17.5cm, width=13cm]{Figure_S5.png}

@@

\rowcolor{lightgray}
\textbf{Practice} & \textbf{Good} & \textbf{Better} & \textbf{Best} \\ \hline

\makecell[l]{Problem \\ definition} & \makecell[l]{Have we clearly stated \\ the ML task? \\ Do we have a priori hypotheses? \\Do we know the predictions \\ a domain expert would \\ make manually?} & \makecell[l]{Do we know the motivation \\ for solving the problem? \\ How much interpretability \\ does the problem need? } & \makecell[l]{Do we know the correlated \\ features? } \\ \hline

\makecell[l]{Classification \\ algorithm} & \makecell[l]{Do we know the \\ candidate algorithms \\ for the ML problem? \\} & \makecell[l]{Do we know our \\ computational resources \\ to fully train each model?} & \makecell[l]{How much interpretability \\ does the problem need? \\ How much each \\ candidate algorithm \\ can provide?} \\ \hline

\makecell[l]{ML pipeline \\ preparation} & \makecell[l]{Did we do \\ cross-validation?} & \makecell[l]{Do we have a held-out \\ test dataset?} & \makecell[l]{Have we tested our model \\ on many different \\  datasets?} \\ \hline

\makecell[l]{Hyperparameter \\ selection} & \makecell[l]{Do we know the different \\ hyperparameters each model \\ can use and why?} & \makecell[l]{Did we use historically \\ effective hyperparameters? } & \makecell[l]{Did we search the \\ full grid space and \\optimized our model? } \\ \hline

\makecell[l]{Model \\ evaluation} & \makecell[l]{Have we chosen an appropriate \\ metric to evaluate \\ predictive performance?} & \makecell[l]{Have we reported \\ the predictive performance \\ on a held-out test data?} & \makecell[l]{Have we provided an average \\ predictive performance \\ of many model runs? } \\ \hline

\makecell[l]{Model \\ interpretation} & \makecell[l]{Do we know if \\ our model is \\ interpretable?} & \makecell[l]{If the model is not \\ interpretable, do we know \\how to  explain it? \\ Have we checked for \\ the effect of correlated \\ features?} & \makecell[l]{Have we generated new \\ hypotheses based on \\ model interpretation \\ to test model results? } \\ \hline

@@

\hypertarget{ref-westcott_opticlust_2017}{}
41. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
Improved Method for Assigning Amplicon-Based Sequence Data to
Operational Taxonomic Units. mSphere \textbf{2}.
doi:\href{https://doi.org/10.1128/mSphereDirect.00073-17}{10.1128/mSphereDirect.00073-17}.

\hypertarget{ref-redwood_stool_2016}{}
42. \textbf{Redwood DG}, \textbf{Asay ED}, \textbf{Blake ID},
\textbf{Sacco PE}, \textbf{Christensen CM}, \textbf{Sacco FD},
\textbf{Tiesinga JJ}, \textbf{Devens ME}, \textbf{Alberts SR},
\textbf{Mahoney DW}, \textbf{Yab TC}, \textbf{Foote PH}, \textbf{Smyrk
TC}, \textbf{Provost EM}, \textbf{Ahlquist DA}. 2016. Stool DNA testing
for screening detection of colorectal neoplasia in alaska native people.
Mayo Clin Proc \textbf{91}:61--70.
doi:\href{https://doi.org/10.1016/j.mayocp.2015.10.008}{10.1016/j.mayocp.2015.10.008}.

\hypertarget{ref-sze_leveraging_2018}{}
43. \textbf{Sze MA}, \textbf{Schloss PD}. 2018. Leveraging existing 16S
rRNA gene surveys to identify reproducible biomarkers in individuals
with colorectal tumors. mBio \textbf{9}:e00630--18.
doi:\href{https://doi.org/10.1128/mBio.00630-18}{10.1128/mBio.00630-18}.

\hypertarget{ref-schloss_introducing_2009}{}
44. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
\textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},
\textbf{Lesniewski RA}, \textbf{Oakley BB}, \textbf{Parks DH},
\textbf{Robinson CJ}, \textbf{Sahl JW}, \textbf{Stres B},
\textbf{Thallinger GG}, \textbf{Van Horn DJ}, \textbf{Weber CF}. 2009.
Introducing mothur: Open-Source, Platform-Independent,
Community-Supported Software for Describing and Comparing Microbial
Communities. ApplEnvironMicrobiol \textbf{75}:7537--7541.

\hypertarget{ref-rognes_vsearch_2016}{}
45. \textbf{Rognes T}, \textbf{Flouri T}, \textbf{Nichols B},
\textbf{Quince C}, \textbf{Mahé F}. 2016. VSEARCH: A versatile open
source tool for metagenomics. PeerJ \textbf{4}:e2584.
doi:\href{https://doi.org/10.7717/peerj.2584}{10.7717/peerj.2584}.

\hypertarget{ref-li_hyperband:_2016}{}
46. \textbf{Li L}, \textbf{Jamieson K}, \textbf{DeSalvo G},
\textbf{Rostamizadeh A}, \textbf{Talwalkar A}. 2016. Hyperband: A novel
bandit-based approach to hyperparameter optimization. arXiv:160306560
{[}cs, stat{]}.

>>>>

Machine learning (ML) modeling of the human microbiome has the potential
to identify microbial biomarkers and aid in diagnosis of many diseases
such as inflammatory bowel disease, diabetes, and colorectal cancer.
Progress has been made towards developing ML models that predict health
outcomes using bacterial abundances, but inconsistent adoption of
training and evaluation methods call the validity of these models into
question. Furthermore, there appears to be a preference by many
researchers to favor increased model complexity over interpretability.
To overcome these challenges, we trained seven models that used fecal
16S rRNA sequence data to predict the presence of colonic screen
relevant neoplasias (SRNs; n=490 patients, 261 controls and 229 cases).
We developed a reusable open-source pipeline to train, validate and
interpret the models. To show the effect of model selection, we assessed
the predictive performance, interpretability, and training time of
L2-regularized logistic regression, L1 and L2-regularized support vector
machines (SVM) with linear and radial basis function kernels, decision
trees, random forest, and gradient boosted trees (XGBoost). The random
forest model performed best at detecting SRNs with an AUROC of 0.695
{[}IQR 0.651-0.739{]} but was slow to train (83.2 h) and not immidiately
interpretable. Despite its simplicity, L2-regularized logistic
regression followed random forest in predictive performance with an
AUROC of 0.680 {[}IQR 0.625-0.735{]}, trained faster (12 min), and
yielded a model that was inherently interpretable. Our analysis
highlights the importance of choosing an ML approach based on the goal
of the study, as the choice will inform expectations of performance and
interpretability.

@@

\textbf{Model selection and pipeline construction}. We established a
reusable ML pipeline for model selection and evaluation, focusing on
seven different commonly used supervised learning algorithms {[}Figure
1{]}.

@@

\textbf{Predictive performance and generalizability of the seven
models.} We evaluated the predictive performance of the seven models to
classify individuals as having normal colons or SRNs {[}Figure 2{]}. The
predictive performance of random forest model was higher than other ML
models with a median 0.695 {[}IQR 0.650-0.739{]}, though not
significantly (p=0.5) (Figure S3). Similarly, L2-regularized logistic
regression, XGBoost, L2-regularized SVM with linear and radial basis
function kernel AUROC values were not significantly different from one
another and had median AUROC values of 0.680 {[}IQR 0.639-0.750{]},
0.679 {[}IQR 0.643-0.746{]}, 0.678 {[}IQR 0.639-0.750{]} and 0.668
{[}IQR 0.639-0.750{]}, respectively. L1-regularized SVM with linear
kernel and decision tree had significantly lower AUROC values than the
other ML models with median AUROC of 0.650 {[}IQR 0.629-0.760{]} and
0.601 {[}IQR 0.636-0.753{]}, respectively {[}Figure 2{]}. Interestingly,
these results demonstrate that the most complex model (XGBoost) did not
have the best performance and that the most interpretable models
(L2-regularized logistic regression and L2-regularized SVM with linear
kernel) performed nearly as well non-linear models.

@@

Second, to analyze non-linear models we interpreted the feature
importance using permutation importance (36). Whereas the absolute
feature weights were determined from the trained models, here we
measured importance using the held-out test data. Permutation importance
analysis is a post hoc explanation of the model, in which we randomly
permuted groups of perfectly correlated features together and other
features individually across the two groups in the held-out test data.
We then calculated how much the predictive performance of the model
(i.e, testing AUROC values) decreased when each OTU or group of OTUs was
randomly permuted. We ranked the OTUs based on how much the median
testing AUROC decreased when it was permuted; the OTU with the largest
decrease ranked highest {[}Figure 4{]}. Among the twenty OTUs with the
largest impact, there was only one OTU (OTU 822) that was shared among
all of the models; however, we found three OTUs (OTUs 58, 110, 367) that
were important in each of the tree-based models. Similarly, the random
forest and XGBoost models, shared four of the most important OTUs (OTUs
2, 12, 361, 477). Permutation analysis results also revealed that with
the exception of the decision tree model, removal of any individual OTU
had minimal impact on model performance. For example, if OTU 367 was
permuted across the samples in the decision tree model, the median AUROC
dropped from 0.601 to 0.525. In contrast, if the same OTU was permuted
in the random forest model, the AUROC only dropped from 0.695 to 0.680
which indicated high degree of collinearity in the dataset. Permutation
analysis allowed us to gauge the importance of each OTU in non-linear
models and partially account for collinearity by grouping correlated
OTUs to determine their impact as a group.

@@

Our framework provides a reproducible structure to investigators wanting
to train, evaluate, and interpret their own ML models to generate
hypotheses regarding which OTUs might be biologically relevant. However,
deploying microbiome-based models to make clinical diagnoses or
predictions is a significantly more challenging and distinct undertaking
(41). For example, we currently lack standardized methods to collect
patient samples, generate sequence data, and report clinical data. We
are also challenged by the practical constraints of OTU-based
approaches. The de novo algorithms commonly in use are slow, require
considerable memory, and result in different OTU assignments as new data
are added (42). Finally, we also need independent validation cohorts to
test the performance of a diagnostic model. To realize the potential for
using ML approaches with microbiome data, it is necessary that we direct
our efforts to overcome these challenges.

@@

\textbf{Data collection and study population.} The original stool
samples described in our analysis were obtained from patients recruited
by Great Lakes-New England Early Detection Research Network (5). Stool
samples were provided by adults who were undergoing a scheduled
screening or surveillance colonoscopy. Participants were recruited from:
Toronto (ON, Canada), Boston (MA, USA), Houston (TX, USA), and Ann Arbor
(MI, USA). Patients' colonic health was visually assessed by colonoscopy
with bowel preparation and tissue histopathology of all resected
lesions. We assigned patients into two classes: those with normal colons
and those with screen relevant neoplasias (SRNs). The normal class
included patients with normal colons or non-advanced adenomas whereas
the SRN class included patients with advanced adenomas or carcinomas
(43). Patients with an adenoma greater than 1 cm, more than three
adenomas of any size, or an adenoma with villous histology were
classified as having advanced adenomas. There were 172 patients with
normal colonoscopies, 198 with adenomas, and 120 with carcinomas. Of the
198 adenomas, 109 were identified as advanced adenomas. Together 261
patients were classified as normal and 229 patients were classified as
having a SRN.

\textbf{16S rRNA gene sequencing data.} Stool samples provided by the
patients were used for 16S rRNA gene sequencing to measure bacterial
population abundances. The sequence data used in our analyses were
originally generated by Baxter et al. (available through NCBI Sequence
Read Archive {[}SRP062005{]}, 2015). The OTU abundance table was
generated by Sze et al (44), who processed the 16S rRNA sequences in
mothur (v1.39.3) using the default quality filtering methods,
identifying and removing chimeric sequences using VSEARCH, and assigning
to OTUs at 97\% similarity using the OptiClust algorithm (42, 45, 46);
(\url{https://github.com/SchlossLab/Sze_CRCMetaAnalysis_mBio_2018/blob/master/data/process/baxter/baxter.0.03.subsample.shared}).
These OTU abundances were the features we used to predict colorectal
health of the patients. There were 6920 OTUs. OTU abundances were
subsampled to the size of the smallest sample and normalized across
samples such that the highest abundance of each OTU would be 1 and
lowest would be 0.

@@

For hyperparameter selection, we started with a granular grid search.
Then we narrowed and fine-tuned the range of each hyperparameter. A grid
search was needed to avoid large variation in prediction performance.
For L2-regularized logistic regression, L1 and L2-regularized SVM with
linear and radial basis function kernels, we tuned the cost
hyperparameter, which controls the regularization strength, where
smaller values specify stronger regularization. For SVM with radial
basis function kernel, we also tuned the sigma hyperparameter which
determines the reach of a single training instance where for a high
value of sigma, the SVM decision boundary will be dependent on the
points that are closest to the decision boundary. For the decision tree
model, we tuned the depth of the tree where the deeper the tree, the
more splits it has. For random forest, we tuned the number of features
to consider when looking for the best tree split. For XGBoost, we tuned
the learning rate and the fraction of samples used for fitting the
individual base learners. Performing a grid search for hyperparameter
selection might not be feasible for when there are more than two
hyperparameters to tune for. In such cases, it is more efficient to use
random search or recently developed tools such as Hyperband to identify
good hyperparameter configurations (47).)

@@

\textbf{Statistical analysis workflow.} Data summaries, statistical
analysis, and data visualizations were performed using R (v.3.5.0) with
the tidyverse package (v.1.2.1). We compared the performance of the
models pairwise by calculating the difference between AUROC values from
the same datasplit (for 100 datasplits). We determined if the models
were significantly different by calculating the empirical p-value (2 x
min(\% of AUROC differences \(\geq\) 0, \% of AUROC differences \(\leq\)
0) for the double tail event (e.g., Figure S3).

@@

\textbf{Acknowlegements.} We thank all the study participants of Great
Lakes-New England Early Detection Research Network. We would like to
thank the members of the Schloss lab for their valuable feedback. Salary
support for M.R came from NIH grant 1R01CA215574. Salary support for
P.D.S. came from NIH grants P30DK034933 and 1R01CA215574.

@@

\textbf{Figure 2. Generalization and classification performance of ML
models using AUROC values of all cross validation and testing
performances.} The median AUROC for diagnosing individuals with SRN
using bacterial abundances was higher than chance (depicted by
horizontal line at 0.50) for all the ML models. The predictive
performance of random forest model was higher than other ML models,s
though not significantly (p=0.5). L2-regularized logistic regression,
XGBoost, L2-regularized SVM with linear and radial basis function kernel
performances were not significantly different from one another. The
boxplot shows quartiles at the box ends and the median as the horizontal
line in the box. The whiskers show the farthest points that were not
outliers. Outliers were defined as those data points that are not within
1.5 times the interquartile ranges. \newpage
\includegraphics[height=17cm, width=12cm]{Figure_3.png}

\textbf{Figure 3. Interpretation of the linear ML models.} The ranks of
absolute feature weights of (A) L1-regularized SVM with linear kernel,
(B) L2-regularized SVM with linear kernel, and (C) L2-regularized
logistic regression, were ranked from highest rank, 1, to lowest rank,
100, for each datasplit. The feature ranks of the 20 highest ranked OTUs
based on their median ranks (median shown in black) are reported here.
OTUs that were associated with classifying a subject as being healthy
had negative signs and were shown in blue. OTUs that were associated
with classifying a subject having an SRN had positive signs and were
shown in red. Some of the same OTUs were identified as important in all
of the linear models. \newpage
\includegraphics{Figure_4.png}

\textbf{Figure 4. Interpretation of the non-linear ML models.} (A) SVM
with radial basis kernel, (B) decision tree, (C) random forest, and (D)
XGBoost feature importances were explained using permutation importance
on the held-out test data set. The gray rectangle and the dashed line
show the IQR range and median of the base testing AUROC without any
permutation. The 20 OTUs that caused the largest decrease in the AUROC
when permuted are reported here. The colors of the box plots represent
the OTUs that were shared among the different models; yellow were OTUs
that were shared among all the non-linear models, salmon were OTUs that
were shared among the tree-based models, green were the OTUs shared
among SVM with radial basis kernel, decision tree and XGBoost, pink were
the OTUs shared among SVM with radial basis kernel and XGBoost only, red
were the OTUs shared among random forest and XGBoost only and blue were
the OTUs shared among decision tree and random forest only. For all of
the tree-based models, a \emph{Peptostreptococcus} species (OTU00367)
had the largest impact on predictive performance. \newpage
\includegraphics{Figure_5.png}

@@

\textbf{Figure S1. Hyperparameter setting performances for linear
models.} (A) L2-regularized logistic regression, (B) L1-regularized SVM
with linear kernel, and (C) L2-regularized SVM with linear kernel mean
cross-validation AUROC values when different hyperparameters were used
in training the model. The stars represent the highest performing
hyperparameter setting for each model. \newpage
\includegraphics[height=30cm, width=15cm]{Figure_S2.png}

\textbf{Figure S2. Hyperparameter setting performances for non-linear
models.} (A) Decision tree, (B) random forest, (C) SVM with radial basis
kernel, and (D) XGBoost mean cross-validation AUROC values when
different hyperparameters were used in training the model. The stars
represent the highest performing hyperparameter setting for the models.
\newpage
\includegraphics[height=17.5cm, width=13cm]{Figure_S3.png}

@@

\textbf{Figure S4. Classification performance of ML models across cross
validation when trained on a subset of the dataset.} (A) L2-regularized
logistic regression and (B) random forest models were trained using the
original study design with 490 subjects and subsets of it with 15, 30,
60, 120, and 245 subjects. The range among the cross-validation AUROC
values within both models at lower sample sizes were much larger than
when the full collection of samples was used to train and validate the
models, but included the ranges observed with the more complete
datasets. \newpage
\includegraphics[height=17.5cm, width=13cm]{Figure_S5.png}

@@

\rowcolor{lightgray}
\textbf{Practice} & \textbf{Poor} & \textbf{Good} & \textbf{Better} \\ \hline

\makecell[l]{Source \\ of data} & \makecell[l]{Input data are not appropriate \\ for intended application. \\ (e.g. input data can be \\ used for diagnosis \\ but not prognosis)} & \makecell[l]{Data are generated according \\ to study design and appropriate \\ for intended application} & \makecell[l]{Data are available \\ to predict intended \\ outcomes and will be \\ available in the future } \\ \hline

\makecell[l]{Study \\ cohort} & \makecell[l]{Test data resampled to \\ remove class imbalance} & \makecell[l]{Test data are reflective \\ of the population which \\ the model will eventually \\ be applied to} & \makecell[l]{Model tested on \\ multiple cohorts \\ with different \\ class balances} \\ \hline

\makecell[l]{Model \\ selection} & \makecell[l]{No justification for \\ classification method} & \makecell[l]{Model choice is justified \\ for intended application} & \makecell[l]{Different modeling \\ choices tested} \\ \hline

\makecell[l]{Model \\ development} & \makecell[l]{No hyperparameter tuning} & \makecell[l]{Tested across multiple \\ hyperparameter settings} & \makecell[l]{Cross-validation performed \\ on training set \\ with hyperparameter \\ grid search} \\ \hline

\makecell[l]{Model \\ evaluation} & \makecell[l]{Performance reported on the data \\ used to train the model} & \makecell[l]{Performance reported on \\ held-out test data} & \makecell[l]{Performance reported on \\ multiple held-out test data } \\ \hline

\makecell[l]{Evaluation \\ metrics} & \makecell[l]{Reported a single metric that is \\ not appropriate for intended \\ application  (e.g. when predicting \\ rare outcome, accuracy metric is \\ not reliable)} & \makecell[l]{Reported multiple metrics} & \makecell[l]{Reported multiple metrics \\ with confidence intervals} \\ \hline

\makecell[l]{Model \\ interpretation} & \makecell[l]{No model interpretation} & \makecell[l]{Follow-up analyses to \\ determine what is driving \\ model performance} & \makecell[l]{Hypotheses on important \\ features are generated \\ and tested} \\ \hline

\makecell[l]{Model \\ specification} & \makecell[l]{Only predictive performance \\ reported} & \makecell[l]{Model described \\ (e.g. feature weights)} & \makecell[l]{How to use the model \\ is explained} \\ \hline
\end{tabular}\newpage

@@

\hypertarget{ref-wiens_no_2019}{}
41. \textbf{Wiens J}, \textbf{Saria S}, \textbf{Sendak M},
\textbf{Ghassemi M}, \textbf{Liu VX}, \textbf{Doshi-Velez F},
\textbf{Jung K}, \textbf{Heller K}, \textbf{Kale D}, \textbf{Saeed M},
\textbf{Ossorio PN}, \textbf{Thadaney-Israni S}, \textbf{Goldenberg A}.
2019. Do no harm: A roadmap for responsible machine learning for health
care. Nat Med \textbf{25}:1337--1340.
doi:\href{https://doi.org/10.1038/s41591-019-0548-6}{10.1038/s41591-019-0548-6}.

\hypertarget{ref-westcott_opticlust_2017}{}
42. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
Improved Method for Assigning Amplicon-Based Sequence Data to
Operational Taxonomic Units. mSphere \textbf{2}.
doi:\href{https://doi.org/10.1128/mSphereDirect.00073-17}{10.1128/mSphereDirect.00073-17}.

\hypertarget{ref-redwood_stool_2016}{}
43. \textbf{Redwood DG}, \textbf{Asay ED}, \textbf{Blake ID},
\textbf{Sacco PE}, \textbf{Christensen CM}, \textbf{Sacco FD},
\textbf{Tiesinga JJ}, \textbf{Devens ME}, \textbf{Alberts SR},
\textbf{Mahoney DW}, \textbf{Yab TC}, \textbf{Foote PH}, \textbf{Smyrk
TC}, \textbf{Provost EM}, \textbf{Ahlquist DA}. 2016. Stool DNA testing
for screening detection of colorectal neoplasia in alaska native people.
Mayo Clin Proc \textbf{91}:61--70.
doi:\href{https://doi.org/10.1016/j.mayocp.2015.10.008}{10.1016/j.mayocp.2015.10.008}.

\hypertarget{ref-sze_leveraging_2018}{}
44. \textbf{Sze MA}, \textbf{Schloss PD}. 2018. Leveraging existing 16S
rRNA gene surveys to identify reproducible biomarkers in individuals
with colorectal tumors. mBio \textbf{9}:e00630--18.
doi:\href{https://doi.org/10.1128/mBio.00630-18}{10.1128/mBio.00630-18}.

\hypertarget{ref-schloss_introducing_2009}{}
45. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
\textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},
\textbf{Lesniewski RA}, \textbf{Oakley BB}, \textbf{Parks DH},
\textbf{Robinson CJ}, \textbf{Sahl JW}, \textbf{Stres B},
\textbf{Thallinger GG}, \textbf{Van Horn DJ}, \textbf{Weber CF}. 2009.
Introducing mothur: Open-Source, Platform-Independent,
Community-Supported Software for Describing and Comparing Microbial
Communities. ApplEnvironMicrobiol \textbf{75}:7537--7541.

\hypertarget{ref-rognes_vsearch_2016}{}
46. \textbf{Rognes T}, \textbf{Flouri T}, \textbf{Nichols B},
\textbf{Quince C}, \textbf{Mahé F}. 2016. VSEARCH: A versatile open
source tool for metagenomics. PeerJ \textbf{4}:e2584.
doi:\href{https://doi.org/10.7717/peerj.2584}{10.7717/peerj.2584}.
